---
title: "Lab 1 key"
author: "Mari Herbst"
date: "1/5/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# attach packages
library(tidyverse)
library(here)
library(sf)
library(tmap)
```

### Read in data

```{r}
sf_trees <- read_csv(here("data","sf_trees","sf_trees.csv"))
```

### Part 1: Wrangling and ggplot review

**Example 1:** Find counts of observation by "legal_status" & wrangle a bit

```{r}
# method 1: group_by() %>% summarize()
sf_trees %>%
  group_by(legal_status) %>%
  summarize(tree_count = n())

# method 2: different way plus a few new functions
top_5_status <- sf_trees %>%
  count(legal_status) %>% # groups by and counts with one function (shortcut for group_by, summarize)
  drop_na(legal_status) %>% # drop NA values in the legal_status column
  rename(tree_count = n) %>% # change column name
  relocate(tree_count) %>% # bring named column to the front spot of the data set
  slice_max(tree_count, n = 5) %>% # only show the top 5 legal statuses by tree count
  arrange(-tree_count) # sets highest to lowest
```

Make a graph of the top 5 legal statuses by tree count

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) + # fct_reorder the legal_status by tree_count instead of in alphabetical order
  geom_col(fill = "darkgreen") +
  labs(x = "legal status",
       y = "tree count") +
  coord_flip() +
  theme_minimal()
```

**Example 2:** Only keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as permitted_data_df

```{r}
permitted_data_df <- sf_trees %>%
  filter(legal_status == "Permitted Site",
         caretaker == "MTA")
```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns "legal_status", "date", "latitude", "longitude", and store as blackwood_acacia_df

```{r}
blackwood_acacia_df <- sf_trees %>%
  filter(str_detect(species, "Blackwood Acacia")) %>% # detects words within a longer string of words so you don't have to type out the whole thing
  select(legal_status, date, lat = latitude, long = longitude) # select desired columns and rename some

# make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = long, y = lat)) +
  geom_point(color = "darkgreen")
```

**Example 4:** Use "tidyr::separate()"

```{r}
sf_trees_sep <- sf_trees %>%
  separate(species, # separate this column
           into = c("spp_scientific", "spp_common"), # into two columns
           sep = " :: ") # break here
```

**Example 5:** Use "tidyr::unite()"

```{r}
ex_5 <- sf_trees %>%
  unite("id_status",
        tree_id, legal_status,
        sep = "_COOL_")
```

### Part 2: Make some maps

**Step 1:** convert lat/long to spatial points, "st_as_sf()"

```{r}
blackwood_acacia_sf <- blackwood_acacia_df %>%
  drop_na(lat, long) %>%
  st_as_sf(coords = c("long", "lat"))

# we need to tell R what the coordinate reference system is
st_crs(blackwood_acacia_sf) <- 4326

ggplot(data = blackwood_acacia_sf) +
  geom_sf(color = "darkgreen")
```

**Step 2:** read in the SF shapefile and add map

```{r}
sf_map <- read_sf(here("data", "sf_map", "tl_2017_06075_roads.shp")) # read in shapefile

sf_map_transform <- st_transform(sf_map, 4326) # transform coordinate ref system to match graph above

ggplot(data = sf_map_transform) +
  geom_sf()
```

**Step 3:** combine the maps!

```{r}
ggplot() +
  geom_sf(data = sf_map,
          size = 0.1,
          color = "darkgrey") +
  geom_sf(data = blackwood_acacia_sf,
          size = 0.5,
          color = "darkgreen") +
  theme_void() +
  labs(title = "Blackwood acacias in SF")
```

### Part 3: Interactive map

```{r}
tmap_mode("view") # set mode to interactive viewing

tm_shape(blackwood_acacia_sf) + # like ggplot
  tm_dots() # like geom_point
```

